A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-30 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: Second Choquet integral figure
This TikZ code reproduces the second sketch for the Choquet integral calculation.

\textbf{Probability of outcomes vs fraction of components.} We saw that classical probability fails in quantum mechanics because pure states are not always mutually exclusive, are not orthogonal. That is why, in quantum mechanics, probability can be defined only once a context, an orthogonal basis, is chosen. Therefore, at a conceptual level, probabilities can only give us a patchwork representation of ensembles. Some authors embrace this view, by characterizing states as the collection of all probability distributions of all possible results, each distribution providing partial information. [CITATION NEEDED] However, there is a different route.

@@

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable as they represent mutually exclusive cases, the measure is additive. In quantum mechanics, only orthogonal states are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states, which equates to the lack of mutual exclusiveness, is responsible for the failure of the additivity of the measure. Note that, since the fraction capacity depends only on the ensemble space, and not also on a chosen ensemble, the fraction capacity essentially characterizes the shape of the ensemble space itself.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{ChoquetCalculation2.jpg}
	\caption{...}
\end{figure}

>>>>

\textbf{Probability of outcomes vs fraction of components.} We saw that classical probability fails in quantum mechanics because pure states are not always mutually exclusive, are not orthogonal. That is why, in quantum mechanics, probability can be defined only once a context, an orthogonal basis, is chosen. Therefore, at a conceptual level, probabilities can only give us a patchwork representation of ensembles. Some author embrace this view, by characterizing states as the collection of all probability distributions of all possible results, each distribution providing partial information. [CITATION NEEDED] However, there is a different route.

@@

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable as they represent mutually exclusive cases, the measure is additive. In quantum mechanics, only orthogonal states are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states, which equates to the lack of mutually exclusiveness, is responsible for the failure of the additivity of the measure. Note that, since the fraction capacity depends only on the ensemble space, and not also on a chosen ensemble, the fraction capacity essentially characterizes the shape of the ensemble space itself.

@@

	% dimensions
	\newcommand{\s}{0.5}
	\pgfmathsetmacro{\radiusats}{sqrt(1 - \s*\s)}
	\pgfmathsetmacro{\stheta}{asin(\s)}
	\pgfmathsetmacro{\minstheta}{asin(-\s)}

	\begin{tikzpicture}[x={(\xx cm, \xy cm)}, y={(\yx cm, \yy cm)}, z={(\zx cm, \zy cm)}, scale=1.2]
		% Bloch sphere (outline drawn in canvas plane)
		\draw (\xy, \yy, \zy) arc[start angle=90, end angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, radius=1 cm] coordinate (A);
%		\draw[dash dot, red] (A) arc[start angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, end angle=-90, radius=1 cm];
		\draw (\xy, \yy, \zy) arc[start angle=90, end angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)}, radius=1 cm] coordinate (B);
%		\draw[dash dot, green] (B) arc[start angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)}, end angle=270, radius=1 cm];
		\draw[dashed] (A) arc[start angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, end angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)-360}, radius=1 cm];

		% special points
		\shadedraw (0, 0, 0) circle (0.03 cm) node[below right] {$\mathsf{o}$};
		\shadedraw (0, \radius, 0) circle (0.03 cm) node[above] {$\mathsf{b}$};
		\shadedraw (0, -\radius, 0) circle (0.03cm) node[below] {$\mathsf{a}$};
		\shadedraw (0, \s, 0) circle (0.03cm) node[left] {$\mathsf{e}$};

		% convex combination
		\draw (0, -\radius, 0) -- (0, \s, 0);
		\draw (2.25*\xx, 2.25*\yx, 2.25*\zx) node {$\ens[o] = \frac{\overline{\ens[o]\ens[a]}}{\overline{\ens[e]\ens[a]}} \ens[e] + \frac{\overline{\ens[o]\ens[e]}}{\overline{\ens[e]\ens[a]}} \ens[a]$};

		% s axis
		\begin{scope}[shift={(-1.5*\xx, -1.5*\yx, -1.5*\zx)}]
			\draw (0, -1, 0) -- (0, 0, 0);
			\draw[->] (0, 0, 0) -- (0, 1, 0) node[below left] {$s$};
		\end{scope}

		% cut
		\begin{scope}[canvas is zx plane at y=\s]
%			\filldraw[color=blue, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
			\draw[blue, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
			\draw[blue, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
		\end{scope}

		% meridians and equator
		\begin{scope}[canvas is yz plane at x=0]
			% foreground above s
			\draw (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, end angle=90-\stheta, radius=1 cm] coordinate (C);
			% foreground below s
			\draw[dashed] (C) arc[start angle=90-\stheta, end angle={atan2(-\ixz, -\ixy)}, radius=1 cm] coordinate (D);
			% background below s
			\draw[loosely dotted] (D) arc[start angle={atan2(-\ixz, -\ixy)}, end angle=\stheta+270, radius=1 cm];
			% background above s
			\draw[densely dotted] (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, end angle=\stheta-90, radius=1 cm];
		\end{scope}
%		\begin{scope}[canvas is xy plane at z=0]
%			\draw[dashed] (\izx, \izy) arc[start angle={atan2(\izy, \izx)}, delta angle=180, radius=1 cm];
%			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, delta angle=180, radius=1 cm];
%		\end{scope}
		\begin{scope}[canvas is xy plane at z=0]
			% foreground above s
			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, end angle=180-\stheta, radius=1 cm] coordinate (E);
			% foreground below s
			\draw[dashed] (E) arc[start angle=180-\stheta, end angle={atan2(\izy, \izx)+360}, radius=1 cm] coordinate (F);
			% background below s
			\draw[loosely dotted] (F) arc[start angle={atan2(\izy, \izx)}, end angle=\stheta, radius=1 cm];
			% background above s
			\draw[densely dotted] (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, end angle=\stheta, radius=1 cm];
		\end{scope}
	\end{tikzpicture}
	\caption{...}
\end{figure}
